---
title: 'Time Series'
author: "OC Data Driven Insights"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  beamer_presentation:
    colortheme: "beaver"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```

# Welcome!

- Instructor: Julian Wolfson, Ph.D., Eric Weber, Ph.D.
- Teaching Assistants: See Moodle
- TA help hours:
    + A446 (Note: not main office)
    + Mon & Tues 11:30-12:30
- TA office hours (Mayo A446) [check Moodle for details]

# Course Website

- https://ay16.moodle.umn.edu/course/view.php?id=10309
    + Syllabus and office hour information
    + Course notes
    + Weekly quizzes
    + Computing information
    + Take-home exam

# Course Communications

- Course announcements come to your x500 (UMN) email account
- You should have already received an email saying no lab this week!

# Books

- Textbook:
    + No formal required text
    + Suggestions: Moore & McCabe (8th Edition)
    + Other suggestions in the syllabus
    
# R Example Import
```{r, eval = FALSE}
path <- "/Users/ericweber/.../drinks.csv"
drinks <- read.csv(path, 
                   stringsAsFactors = FALSE, 
                   header = TRUE)
```

# 5 Big Ideas

- Why is randomness important?
- What is variability?
- Why is variability important?
- Sources of variability
- Concrete examples of random variables
